Probabilistic Catalogs for Crowded Stellar Fields

### 2012-11-23 paper/ms.tex
Removed stray [ht!]
Comma
Coordinate range of data

Bayesian object detection (as this problem is sometimes called) has been
implemented both inside and outside of astronomy
\citep[e.g.][]{object, 2011MNRAS.415.3462F}. However, the
\citet{2011MNRAS.415.3462F} approach makes the assumption of a
known number of objects $N$.
This assumption is required for the
{\tt MultiNest} sampler \citep{multinest} to be applicable.
Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if an unknown-$N$ model had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N \in \{0, 1, 2, ... \}$ is an unknown quantity to be
inferred from the data.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}. Other statistical
methods have also been used to model crowded fields
\citep[e.g. maximum likelihood][]{irwin}.
However, maximum likelihood is not completely appropriate in flexible models
because it may lead to overfitting. In this situation overfitting would result
in more stars being added to the model to explain small positive fluctuations
in the image which are actually due to noise. Various other techniques have
also been proposed in the literature \citep[e.g.][]{2002MNRAS.335...73M, 2007A&A...461..373M, 2009PASP..121.1429Z}.

@@

\subsection{The Prior Distribution}
The prior distribution for the number of stars $N$ is assigned to be uniform
between 0 and some maximum number $N_{\rm max}$. The extent of the image is
assumed to be from $x=x_{\rm min}$ to $x=x_{\rm max}$ and from
$y=y_{\rm min}$ to $y=y_{\rm max}$ in arbitrary units, and the
positions of the stars are assigned independent uniform priors:
\begin{eqnarray}
x_i &\sim& \textnormal{Uniform}(1.1 x_{\rm min}, 1.1 x_{\rm max}) \\
y_i &\sim& \textnormal{Uniform}(1.1 y_{\rm min}, 1.1 y_{\rm max})
\end{eqnarray}
The stars are allowed to be slightly outside of the observed image because the
PSF can scatter light from these stars into the image. Our model for the
luminosity function is a broken power-law with four
free parameters:
\begin{eqnarray}
\beta = \{h_1, h_2, \alpha_1, \alpha_2\}.
\end{eqnarray}
where $h_1$ is a lower flux limit, $h_2$ is a break-point, $\alpha_1$ is
the slope of the distribution between $h_1$ and $h_2$, and $\alpha_2$ is the
slope of the distribution above $h_2$. For details on the broken power-law
model, see Appendix~\ref{power_law}. The prior distribution on $h_1$, $h_2$,
$\alpha_1$, and $\alpha_2$ is assigned to be:
\begin{eqnarray}
\ln h_1 &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^3)) \\
\ln h_2 &\sim& \textnormal{Uniform}(\ln(h_1), \ln(h_1) + 2.3) \\
\alpha_1 &\sim& \textnormal{Uniform}(1, 5) \\
\alpha_2 &\sim& \textnormal{Uniform}(1, 5).
\end{eqnarray}
These priors express vague prior knowledge about $\alpha_1$ and $\alpha_2$ in
addition to vague prior knowledge about $h_1$ and $h_2$ apart from the fact
that the flux units are not extreme and that $h_2$ should be no more than an
order of magnitude greater than $h_1$.

@@

\section{Broken Power-Law Distribution}\label{power_law}
The broken power-law distribution is based on a straightforward extension
to a simple power-law distribution (also known as a Pareto distribution,
particularly in the statistics literature). The power-law distribution for a
variable $x$ (given a lower cutoff $x=h$ and a slope $\alpha$) is defined by:
\begin{eqnarray}
p(x) &\propto&
\left\{
\begin{array}{lcl}
0, & & x < h \\
x^{-\alpha - 1} & & x \geq h.
\end{array}
\right.
\end{eqnarray}
In contrast, the broken power-law distribution for a variable $x$ is defined by
a lower cutoff $x=h_1$, two slopes $\{\alpha_1, \alpha_2\}$ and a break point
$x=h_2$:
\begin{eqnarray}
p(x) &\propto&
\left\{
\begin{array}{lcrl}
0, & & x < h_1 & \\
x^{-\alpha_1 - 1} & & h_1 \leq x \leq h_2 & \\
x^{-\alpha_2 - 1} & & x > h_2.
\end{array}
\right.
\end{eqnarray}
The free parameters of the broken power-law are:
\begin{eqnarray}
\beta = \{h_1, h_2, \alpha_1, \alpha_2\}.
\end{eqnarray}
With normalising terms included, the proportionality becomes an equality:
\begin{eqnarray}
p(x) &=&
\left\{
\begin{array}{lcr}
0, & & x < h_1 \\
Z_1^{-1}x^{-\alpha_1 - 1}, & & h_1 \leq x \leq h_2 \\
Z_2^{-1}x^{-\alpha_2 - 1}, & & x > h_2.
\end{array}
\right.
\end{eqnarray}
Two conditions will be used to determine the normalizers $Z_1$ and $Z_2$.
Firstly, the probability density function (PDF) should be continuous at $x=h_2$:
\begin{eqnarray}
Z_1^{-1}h_2^{-\alpha_1 - 1} &=& Z_2^{-1}h_2^{-\alpha_2 - 1}\\
\implies
Z_2 &=& Z_1h_2^{\alpha_1-\alpha_2}
\end{eqnarray}
The second condition is that the total probability must be 1:
\begin{eqnarray}
\int_{h_1}^{h_2} Z_1^{-1} x^{-\alpha_1 - 1} \, dx
+
\int_{h_2}^\infty Z_2^{-1} x^{-\alpha_2 - 1} \, dx
&=& 1 \\
Z_1^{-1}\alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
Z_2^{-1}\alpha_2^{-1}h_2^{-\alpha_2}
&=& 1 \\
Z_1^{-1}\alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
Z_1^{-1}h_2^{\alpha_2-\alpha_1}\alpha_2^{-1}h_2^{-\alpha_2}
&=& 1
\end{eqnarray}
\begin{eqnarray}
\implies
Z_1 &=& \alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
h_2^{-\alpha_1}\alpha_2^{-1}.
\end{eqnarray}
The cumulative distribution (CDF) is a useful property of a probability
distribution and is given by the antiderivative of the PDF:
\begin{eqnarray}
P(X \leq x) = F(x) &=&
\left\{
\begin{array}{lcr}
0, & & x < h_1 \\
(Z_1\alpha_1)^{-1}\left(h_1^{-\alpha_1} - x^{-\alpha_1}\right), & & h_1 \leq x \leq h_2 \\
1 - (Z_2\alpha_2)^{-1}x^{-\alpha_2}, & & x > h_2.
\end{array}
\right.
\end{eqnarray}
The inverse of the CDF is also useful and is given by:
\begin{eqnarray}
F^{-1}(u) &=&
\left\{
\begin{array}{lcr}
\left[h_1^{-\alpha_1} - uZ_1\alpha_1\right]^{-1/\alpha_1}, & & 0 < u < 1 - (Z_2\alpha_2)^{-1}h_2^{-\alpha_2}\\
\left[Z_2\alpha_2(1-u)\right]^{-1/\alpha_2},& & 1 - (Z_2\alpha_2)^{-1}h_2^{-\alpha_2}< u < 1.
\end{array}
\right.
\end{eqnarray}
An example of a broken power-law distribution is shown in Figure~\ref{fig:powerlaw}.
\begin{figure}[ht!][ht!]
\begin{center}
\includegraphics[width=\textwidth]{Figures/broken.eps}
\caption{A broken power-law distribution. The parameter values for this
particular PDF were
$\{h_1, h_2, \alpha_1, \alpha_2\} = \{0.3, 0.6, 1.1, 2\}$, i.e. the same
parameter values used to make the simulated data.
\label{fig:powerlaw}}
\end{center}
\end{figure}

>>>>

Bayesian object detection (as this problem is sometimes called) has been
implemented both inside and outside of astronomy
\citep[e.g.][]{object, 2011MNRAS.415.3462F}. However, the
\citet{2011MNRAS.415.3462F} approach makes the assumption of a
known number of objects $N$.
This assumption is required for the
{\tt MultiNest} sampler \citep{multinest} to be applicable.
Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if an unknown-$N$ model had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N \in \{0, 1, 2, ... \}$ is an unknown quantity to be
inferred from the data.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}. Other statistical
methods have also been used to model crowded fields
\citep[e.g. maximum likelihood,][]{irwin}.
However, maximum likelihood is not completely appropriate in flexible models
because it may lead to overfitting. In this situation overfitting would result
in more stars being added to the model to explain small positive fluctuations
in the image which are actually due to noise. Various other techniques have
also been proposed in the literature \citep[e.g.][]{2002MNRAS.335...73M, 2007A&A...461..373M, 2009PASP..121.1429Z}.

@@

\subsection{The Prior Distribution}
The prior distribution for the number of stars $N$ is assigned to be uniform
between 0 and some maximum number $N_{\rm max}$. The extent of the image is
assumed to be from $x=x_{\rm min}$ to $x=x_{\rm max}=x_{\rm min} + x_{\rm range}$ and from
$y=y_{\rm min}$ to $y=y_{\rm max}=y_{\rm min} + y_{\rm range}$ in arbitrary units, and the
positions of the stars are assigned independent uniform priors:
\begin{eqnarray}
x_i &\sim& \textnormal{Uniform}(x_{\rm min} - 0.05x_{\rm range}, x_{\rm max} + 0.05x_{\rm range}) \\
y_i &\sim& \textnormal{Uniform}(y_{\rm min} - 0.05y_{\rm range}, y_{\rm max} + 0.05y_{\rm range})
\end{eqnarray}
The stars are allowed to be slightly outside of the observed image because the
PSF can scatter light from these stars into the image. Our model for the
luminosity function is a broken power-law with four
free parameters:
\begin{eqnarray}
\beta = \{h_1, h_2, \alpha_1, \alpha_2\}.
\end{eqnarray}
where $h_1$ is a lower flux limit, $h_2$ is a break-point, $\alpha_1$ is
the slope of the distribution between $h_1$ and $h_2$, and $\alpha_2$ is the
slope of the distribution above $h_2$. For details on the broken power-law
model, see Appendix~\ref{power_law}. The prior distribution on $h_1$, $h_2$,
$\alpha_1$, and $\alpha_2$ is assigned to be:
\begin{eqnarray}
\ln h_1 &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^3)) \\
\ln h_2 &\sim& \textnormal{Uniform}(\ln(h_1), \ln(h_1) + 2.3) \\
\alpha_1 &\sim& \textnormal{Uniform}(1, 5) \\
\alpha_2 &\sim& \textnormal{Uniform}(1, 5).
\end{eqnarray}
These priors express vague prior knowledge about $\alpha_1$ and $\alpha_2$ in
addition to vague prior knowledge about $h_1$ and $h_2$ apart from the fact
that the flux units are not extreme and that $h_2$ should be no more than an
order of magnitude greater than $h_1$.

@@

\section{Broken Power-Law Distribution}\label{power_law}
The broken power-law distribution is based on a straightforward extension
to a simple power-law distribution (also known as a Pareto distribution,
particularly in the statistics literature). The power-law distribution for a
variable $x$ (given a lower cutoff $x=h$ and a slope $\alpha$) is defined by:
\begin{eqnarray}
p(x) &\propto&
\left\{
\begin{array}{lcl}
0, & & x < h \\
x^{-\alpha - 1} & & x \geq h.
\end{array}
\right.
\end{eqnarray}
In contrast, the broken power-law distribution for a variable $x$ is defined by
a lower cutoff $x=h_1$, two slopes $\{\alpha_1, \alpha_2\}$ and a break point
$x=h_2$:
\begin{eqnarray}
p(x) &\propto&
\left\{
\begin{array}{lcrl}
0, & & x < h_1 & \\
x^{-\alpha_1 - 1} & & h_1 \leq x \leq h_2 & \\
x^{-\alpha_2 - 1} & & x > h_2.
\end{array}
\right.
\end{eqnarray}
The free parameters of the broken power-law are:
\begin{eqnarray}
\beta = \{h_1, h_2, \alpha_1, \alpha_2\}.
\end{eqnarray}
With normalising terms included, the proportionality becomes an equality:
\begin{eqnarray}
p(x) &=&
\left\{
\begin{array}{lcr}
0, & & x < h_1 \\
Z_1^{-1}x^{-\alpha_1 - 1}, & & h_1 \leq x \leq h_2 \\
Z_2^{-1}x^{-\alpha_2 - 1}, & & x > h_2.
\end{array}
\right.
\end{eqnarray}
Two conditions will be used to determine the normalizers $Z_1$ and $Z_2$.
Firstly, the probability density function (PDF) should be continuous at $x=h_2$:
\begin{eqnarray}
Z_1^{-1}h_2^{-\alpha_1 - 1} &=& Z_2^{-1}h_2^{-\alpha_2 - 1}\\
\implies
Z_2 &=& Z_1h_2^{\alpha_1-\alpha_2}
\end{eqnarray}
The second condition is that the total probability must be 1:
\begin{eqnarray}
\int_{h_1}^{h_2} Z_1^{-1} x^{-\alpha_1 - 1} \, dx
+
\int_{h_2}^\infty Z_2^{-1} x^{-\alpha_2 - 1} \, dx
&=& 1 \\
Z_1^{-1}\alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
Z_2^{-1}\alpha_2^{-1}h_2^{-\alpha_2}
&=& 1 \\
Z_1^{-1}\alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
Z_1^{-1}h_2^{\alpha_2-\alpha_1}\alpha_2^{-1}h_2^{-\alpha_2}
&=& 1
\end{eqnarray}
\begin{eqnarray}
\implies
Z_1 &=& \alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
h_2^{-\alpha_1}\alpha_2^{-1}.
\end{eqnarray}
The cumulative distribution (CDF) is a useful property of a probability
distribution and is given by the antiderivative of the PDF:
\begin{eqnarray}
P(X \leq x) = F(x) &=&
\left\{
\begin{array}{lcr}
0, & & x < h_1 \\
(Z_1\alpha_1)^{-1}\left(h_1^{-\alpha_1} - x^{-\alpha_1}\right), & & h_1 \leq x \leq h_2 \\
1 - (Z_2\alpha_2)^{-1}x^{-\alpha_2}, & & x > h_2.
\end{array}
\right.
\end{eqnarray}
The inverse of the CDF is also useful and is given by:
\begin{eqnarray}
F^{-1}(u) &=&
\left\{
\begin{array}{lcr}
\left[h_1^{-\alpha_1} - uZ_1\alpha_1\right]^{-1/\alpha_1}, & & 0 < u < 1 - (Z_2\alpha_2)^{-1}h_2^{-\alpha_2}\\
\left[Z_2\alpha_2(1-u)\right]^{-1/\alpha_2},& & 1 - (Z_2\alpha_2)^{-1}h_2^{-\alpha_2}< u < 1.
\end{array}
\right.
\end{eqnarray}
An example of a broken power-law distribution is shown in Figure~\ref{fig:powerlaw}.
\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{Figures/broken.eps}
\caption{A broken power-law distribution. The parameter values for this
particular PDF were
$\{h_1, h_2, \alpha_1, \alpha_2\} = \{0.3, 0.6, 1.1, 2\}$, i.e. the same
parameter values used to make the simulated data.
\label{fig:powerlaw}}
\end{center}
\end{figure}
